Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-06 tagless/tagless-final.tex
Revise abstract (and small changes elsewhere)

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}
\usepackage{mdwtab}
\usepackage{dblfloatfix}

@@

\begin{abstract}
We have built a family of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has been an open problem.
Our family of interpreters contains 
an evaluator, a (byte-)compiler and staged evaluator, a partial
evaluator, a term-size evaluator, and call-by-name and
call-by-value CPS transformers.

The main tool is to take functions seriously: we encode the
de Bruijn or higher-order abstract syntax  of object terms 
not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
\oleg{Or: we expose the co-algebraic structure of the object language.}
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), available with ML functors and
Haskell98 constructor classes. This lets us encode object terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation
never gets stuck.

The encoding of a self-interpreter clarifies our technique 
\oleg{and rank-2 polymorphism?} 
as the
interpretation \oleg{evaluation?} of a term plugged into the context
of let-polymorphic definitions.
\oleg{Discuss on Fri, along with the colclusions.}
\end{abstract}

@@

A popular way to define and implement a programming language is to embed it in
another \citep{reynolds-definitional}.  Embedding means to represent
terms and values of the \emph{object language} as terms and values in the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \citep{hudak-building}.
If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser or type checker reading from a text string.
We leave aside the solved problem of writing such a parser or type checker,
whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

Our code ensures statically that an object term is well-typed and
closed, yet is surprisingly simple and obvious in hindsight.  We stress
that until now it has been considered impossible to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

Our contributions are as follows.
\begin{enumerate}
\item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
  compile, and partially evaluate a typed higher-order object language
   in a typed metalanguage.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.
\item We use the type system of the metalanguage
    to check statically for well-typed and closed object programs.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS transformation.
\item We point a clear way to extend the object language with more features
  such as state.
\item We describe an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out harder than expected.
\end{enumerate}

The structure of our paper is as follows. In the next Section, we
define the OCaml and Haskell interfaces of the interpreter for our
language and describe two instances, evaluating a term to a
metalanguage value and computing term's size. In
Section~\ref{compiler}, we stage the evaluator, obtaining the tagless
compiler. Section~\ref{PE} describes partial evaluation, in
particular, the problem posed by the need to delay the 
binding-time analysis of term abstractions and the apparent 
need for a typecase. We then show two solutions. The first, imperfect
one, uses GADT. The better and more portable solution bakes in the
typecase idiom into the interpreter interface, eliminating the need for
advanced type systems and permitting implementations in Haskell98 or OCaml.
The discussion Section~\ref{discussion} presents, inter alia, CBN CPS
interpreter and the CBV CPS transformer. We then describe 
self-interpretation (Section~\ref{selfinterp}),
discuss related work (Section~\ref{related}) and then conclude.

>>>>

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
\usepackage{mdwlist}
\usepackage{mdwtab}
\usepackage{dblfloatfix}

@@

\begin{abstract}
We have built the first family of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell or ML) that
require no dependent
types, generalized algebraic data types, or
postprocessing to eliminate tags.
The interpretations include
an evaluator, a compiler (or staged evaluator), a partial
evaluator, and call-by-name and
call-by-value CPS transformers.

Our main idea is to encode
de Bruijn or higher-order abstract syntax
using cogen functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

Our encoding of an object term abstracts over the various ways to
interpret it yet statically assures that the interpreters never get
stuck.  To achieve self\hyp interpretation and show Jones\hyp
optimality, we relate this exemplar of higher-rank and higher-kind
polymorphism (provided by ML functors and Haskell~98 constructor
classes) to plugging a term into a context of let\hyp polymorphic
bindings.
\end{abstract}

@@

A popular way to define and implement a language is to embed it in
another \citep{reynolds-definitional}.  Embedding means to represent
terms and values of the \emph{object language} as terms and values in the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \citep{hudak-building}.
If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser\slash type\hyp checker reading from a text string.
We leave aside the solved problem of writing such a parser\slash type\hyp checker,
whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

Our code ensures statically that an object term is well-typed and
closed, yet is surprisingly simple and obvious in hindsight.  We stress
that it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
  compile, and partially evaluate a typed higher-order object language
   in a typed metalanguage.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.
\item We use the type system of the metalanguage
    to check statically for well-typed and closed object programs.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS transformation.
\item We point a clear way to extend the object language with more features
  such as state.
\item We describe an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}

The structure of our paper is as follows. In~\S\ref{language}, we
define the OCaml and Haskell interfaces of the interpreter for our
language and describe two instances, evaluating a term to a
metalanguage value and computing term's size.
In~\S\ref{compiler}, we stage the evaluator, obtaining the tagless
compiler. \S\ref{PE} describes partial evaluation, in
particular, the problem posed by the need to delay the 
binding-time analysis of term abstractions and the apparent 
need for a typecase. We then show two solutions. The first, imperfect
one, uses GADT. The better and more portable solution bakes in the
typecase idiom into the interpreter interface, eliminating the need for
advanced type systems and permitting implementations in Haskell98 or OCaml.
The discussion \S\ref{discussion} presents, inter alia, CBN CPS
interpreter and the CBV CPS transformer. We then describe 
self-interpretation (\S\ref{selfinterp}),
discuss related work (\S\ref{related}) and then conclude.
